title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/home/tucker/Documents/GitHub/BEwebinar/Data')
```

# Lesson 9-5: Sample Autocovariance



## Remark 9.5.5. The Sample ACVF

- The so-called *sample autocovariance function* (sample ACVF) is obtained by replacing $\mu$ by $\overline{X}$ in $\overline{\gamma} (h)$, and rescaling to guarantee a positive definite sequence:
\[
 \widehat{\gamma} (h) = \frac{1}{n} \sum_{t=1}^{n-h} (X_{t+h} - \overline{X})
  (X_t - \overline{X})
\]
 for $h \geq 0$. 
- Under $m$-dependence, Corollary 9.5.8 establishes a central limit theorem for the sample ACVF:
  \[
   \sqrt{n} ( \widehat{\gamma} (h)  - \gamma (h)) \Rightarrow \mathcal{N} (0, \tau^2_{\infty}).
\]

## Exercise 9.28. Sample ACVF of ARMA(1,2) Process

- We simulate a Gaussian ARMA(1,2) process of length $n=200$.

```{r}
armapq.sim <- function(n,burn,ar.coefs,ma.coefs,innovar)
{
	p <- length(ar.coefs)
	q <- length(ma.coefs)
	z <- rnorm(n+burn+p+q,sd=sqrt(innovar))
	x <- filter(z,c(1,ma.coefs),method="convolution",sides=1)
	x <- x[(q+1):(q+n+burn+p)]
	y <- x[1:p]
	for(t in (p+1):(p+n+burn))
	{
		next.y <- sum(ar.coefs*y[(t-1):(t-p)]) + x[t]
		y <- c(y,next.y)
	}	
	y <- y[(p+burn+1):(p+burn+n)]
	return(y)
}

n <- 200
phi1 <- .5
theta1 <- 5/6
theta2 <- 1/6
x.sim <- armapq.sim(n,500,phi1,c(theta1,theta2),1)
```

- We construct and plot the sample acvf.

```{r}
y.sim <- x.sim - mean(x.sim)
x.acf <- mean(y.sim^2)
for(h in 1:20)
{
	x.acf <- c(x.acf,sum(y.sim[1:(n-h)]*y.sim[(h+1):n])/n)
}
plot(ts(x.acf,start=0,frequency=1),xlab="Lag",ylab="Sample Autocovariance",type="h")
```
