---
title: 'Time Series: A First Course with Bootstrap Sampler'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/TimeSeriesR/Data')
```

# Lesson 8: Linear Filters

- A *linear filter* maps an input time series $\{ X_t \}$ to an output time series $\{ Y_t \}$ by taking a linear combination of past, present, and future observations:
\[
  Y_t = \sum_{k \in {\mathbb Z}} \psi_k \, X_{t-k}.
\]
 The *filter coefficients* (or *filter weights*) are $\{ \psi_k \}$. Note the convention that $\psi_k$ weights an observation occuring $k$ time points in the past (viewing time $t$ as the present). 
 
## Example 3.2.5.  Simple Moving Average

- In nonparametric regression we may estimate a slowly-changing mean via averaging over neighboring values, e.g.,
\[
   \frac{1}{2m+1} \sum_{k = -m}^m   X_{t-k}.
\]
 This weights the past $m$ and the future $m$ observations equally, and is an example of a linear filter.  It also called a *moving average*, because the observations are averaged (over a window of $2m+1$ time points) in a way that moves over the time series. A *simple moving average* has equal weights (a general moving average could have unequal weights).
 
## Example 3.3.2. Smoothers Applied to Gasoline Sales

- Linear filters that suppress oscillations and reveal long-term trends are called *smoothers*. A simple moving average is an example of a smoother.
- We apply a simple moving average with $m=10$ to the logged seasonally adjusted gasoline series.

```{r}
gassa <- read.table("GasSA_2-11-13.dat")
gassa.log <- ts(log(gassa),start=1992,frequency=12)

h <- 10
simple.ma <- rep(1,2*h+1)/(2*h+1)
gas.trend <- filter(gassa.log,simple.ma,method="convolution",sides=2)
gas.trend <- ts(gas.trend,start=1992,frequency=12)
plot(ts(gassa.log,start=1992,frequency=12),col=1,ylab="",xlab="Year",lwd=2)
lines(ts(gas.trend,start=1992,frequency=12),col=2,lwd=2)

### Movie
movie <- TRUE
delay <- 0
flex <- 0
siglen <- length(gas.trend) - 2*h
range <- seq(1,siglen,1)
if(movie) {
for(t in range)
{
	Sys.sleep(delay)
	filterweight <- c(rep(NA,t-1),simple.ma + gas.trend[h+t] + flex,
		rep(NA,siglen-t))
	plot(ts(gassa.log,start=1992,frequency=12),col=1,ylab="",xlab="Year",lwd=2)
	lines(ts(gas.trend[1:(h+t)],start=1992,frequency=12),col=2,lwd=2)
	lines(ts(filterweight,start=1992,frequency=12),col="#00FF0090",lwd=2)
} }
```


## Paradigm 3.3.3.  Difference Filter

- To suppress long-term dynamics, we can *difference*: 
\[
   Y_t = X_t - X_{t-1}.
\]
 This is a filter with weights $\psi_0 = 1$, $\psi_1 = -1$, and zero otherwise. It is called the *differencing filter*. This filter reduces polynomials in time by one degree; lines are reduced to constants.
 
## Example 3.3.4. Differencing Applied to Gasoline Sales

- We apply differencing to logged seasonally adjusted gasoline sales. The result can be interpreted as a *growth rate*. 

```{r}
gas.diff <- diff(gassa.log)
par(oma=c(2,0,0,0),mar=c(2,4,2,2)+0.1,mfrow=c(2,1),cex.lab=.8)
plot(gassa.log,xlab="",ylab="Log Gas  Sales",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(gas.diff,xlab="",ylab = "Gas Growth Rate",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
mtext(text="Year",side=1,line=1,outer=TRUE)
```

## The Backward Shift

- We define the *backward shift* (or *lag*) operator $B$, which shifts a time series back one time unit. This is a linear filter that lags time by one unit:
\[
    Y_t = X_{t-1}.
\]
 So the filter weights are $\psi_1 = 1$ and zero otherwise. Informally we write $B X_t = X_{t-1}$.
 
## Powers of Backward Shift

- Then $X_{t-k} = B^k X_t$, and a general filter is expressed as
\[
  Y_t = \sum_{k \in {\mathbb Z}} \psi_k \, X_{t-k}
   =  \sum_{k \in {\mathbb Z}} \psi_k B^k \, X_{t},
\]
 so we write the filter as  $\sum_{k \in {\mathbb Z}} \psi_k B^k$. Call this $\Psi (B)$. (Mathematically, $\Psi (z)$ is a Laurent series.)
 
### Simple Moving Average Filter

- Expressed as $\Psi (B) = {(2m+1)}^{-1} \sum_{k = -m}^m B^k$.

## Differencing Filter

- Expressed as $\Psi (B) = 1 -B$.
  