---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 6-2: Filtering in Frequency Domain

- Filters extract (or suppress) features of interest from a time series.
- Using Corollary 6.1.9, we can see in frequency domain how extraction and suppression occurs.

## Example 6.2.1. Business Cycle in Housing Starts.

- Example 3.6.13 decomposed West Housing Starts into Trend, Seasonal, and Irregular.

```{r}
hpsa <- function(n,period,q,r)
{
	# hpsa
	#     gives an HP filter for seasonal data
	#	presumes trend+seas+irreg structure
	#		trend is integrated rw
	#		seas is seasonal rw
	#		irreg is wn
	#	q is snr for trend to irreg
	#	r is snr for seas to irreg

# define trend differencing matrix

delta.mat <- diag(n)
temp.mat <- 0*diag(n)
temp.mat[-1,-n] <- -2*diag(n-1)
delta.mat <- delta.mat + temp.mat
temp.mat <- 0*diag(n)
temp.mat[c(-1,-2),c(-n,-n+1)] <- 1*diag(n-2)
delta.mat <- delta.mat + temp.mat
diff.mat <- delta.mat[3:n,]

# define seasonal differencing matrix

delta.mat <- diag(n)
temp.mat <- 0*diag(n)
inds <- 0
for(t in 1:(period-1))
{
	temp.mat <- 0*diag(n)
	temp.mat[-(1+inds),-(n-inds)] <- 1*diag(n-t)
	delta.mat <- delta.mat + temp.mat
	inds <- c(inds,t)
}
sum.mat <- delta.mat[period:n,]

# define two-comp sig ex matrices

#trend.mat <- solve(diag(n) + t(diff.mat) %*% diff.mat/q)
#seas.mat <- solve(diag(n) + t(sum.mat) %*% sum.mat/r)
trend.mat <- diag(n) - t(diff.mat) %*% solve(q*diag(n-2) + diff.mat %*% 
	t(diff.mat)) %*% diff.mat
seas.mat <- diag(n) - t(sum.mat) %*% solve(r*diag(n-period+1) + sum.mat %*% 
	t(sum.mat)) %*% sum.mat

# define three-comp sig ex matrices

trend.filter <- solve(diag(n) - trend.mat %*% seas.mat) %*%
	trend.mat %*% (diag(n) - seas.mat)
seas.filter <- solve(diag(n) - seas.mat %*% trend.mat) %*%
	seas.mat %*% (diag(n) - trend.mat)
irreg.filter <- diag(n) - (trend.filter + seas.filter)

filters <- list(trend.filter,seas.filter,irreg.filter)
return(filters)
}

Wstarts <- read.table("Wstarts.b1",skip=2)[,2]
Wstarts <- ts(Wstarts,start = 1964,frequency=12)
n <- length(Wstarts)
q <- .0001
r <- 1
hp.filters <- hpsa(n,12,q,r)

wstarts.trend <- ts(hp.filters[[1]] %*% Wstarts,start=1964,frequency=12)
wstarts.seas <- ts(hp.filters[[2]] %*% Wstarts,start=1964,frequency=12)
wstarts.irreg <- ts(hp.filters[[3]] %*% Wstarts,start=1964,frequency=12)

par(oma=c(2,0,0,0),mar=c(2,4,2,2)+0.1,mfrow=c(4,1),cex.lab=.8)
plot(Wstarts, ylab="West Starts",xlab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(wstarts.trend,xlab="",ylab = "Trend",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(wstarts.seas,xlab="",ylab = "Seasonal",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(wstarts.irreg,xlab="",ylab = "Irregular",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
mtext(text="Year",side=1,line=1,outer=TRUE)
```

- We fit (e.g. via ordinary least squares) an AR(26) to the Irregular component, 
and plot the log spectral density.
- The units are in terms of $2 \pi/12$, so the x-axis numbers represent multiples 
of $\pi/6$.
- There are vertical bands for frequency between $.5$ and $.1$, corresponding to period
between 2 and 10 years. This is the *business cycle* range.

```{r}
ar.fit <- spec.ar(ts(wstarts.irreg,frequency=12),plot=FALSE)
plot(ts(log(ar.fit$spec),start=0,frequency=500/6),xlab="Frequency per Year",
	ylab="Log Spectrum",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
abline(v=.5,lty=3)
abline(v=.1,lty=3)
```

## Remark 6.2.2. Spectral Peaks and Oscillation Frequencies

- Higher values of the spectral density correspond to frequencies with more variability.
- A peak in the spectral density at a frequency $\lambda$ corresponds to an
oscillation, or cyclical effect in the process.

## Fact 6.2.5. Suppression and Extraction

- Suppose $\{ X_t \}$ is filtered with $\psi (B)$, so that $Y_t = \psi (B) X_t$.
- If $\psi (e^{-i \lambda}) = 0$, then $f_y (\lambda) = 0$, and $\lambda$ is *suppressed*. 
The set of such frequencies is the *stop-band*.
- If $\psi (e^{-i \lambda}) = 1$, then $f_y (\lambda) = f_x (\lambda)$, and $\lambda$ is *extracted*. 
The set of such frequencies is the *pass-band*.

## Exercise 6.22. The Ideal Low-Pass Filter

- The Ideal Low-pass is defined with frequency response function
\[
 \psi (e^{-i \lambda}) = \begin{cases}
 1 \qquad \mbox{if} \; |\lambda | \leq \mu \\ 0 \qquad \mbox{else}.
 \end{cases}
\]
- We plot with cut-off $\mu = \pi/5$.

```{r}
mu <- pi/5
mesh <- 1000
lambda <- pi*seq(0,mesh)/mesh
psi.frf <- rep(0,mesh+1)
psi.frf[lambda <= mu] <- 1
plot(ts(psi.frf,start=0,frequency=mesh),ylab="",xlab="Cycles")
```

- It is hard to implement, since the filter coefficients decay slowly (and hence
truncation is expensive).

## Exercise 6.23. The Ideal Band-Pass Filter

- The Ideal Band-pass is defined with frequency response function
\[
 \psi (e^{-i \lambda}) = \begin{cases}
 1 \qquad \mbox{if} \; \mu_1 < |\lambda | \leq \mu_2 \\ 0 \qquad \mbox{else}.
 \end{cases}
\]
- We plot with cut-offs $\mu_1 = \pi/60$ and $\mu_2 = \pi/12$.

```{r}
mu1 <- pi/60
mu2 <- pi/12
mesh <- 1000
lambda <- pi*seq(0,mesh)/mesh
psi.frf <- rep(0,mesh+1)
psi.frf[lambda <= mu2] <- 1
psi.frf[lambda <= mu1] <- 0
plot(ts(psi.frf,start=0,frequency=mesh),ylab="",xlab="Cycles")
```

## Example 6.2.7. The Hodrick-Prescott Filter

- Proposed by Whitaker in 1927, but known as Hodrick-Prescott, the filter does trend
extraction.
- The frequency response function resembles an ideal low-pass.
- For the decomposition of West Housing Starts, we used a modified Hodrick-Prescott
that is adapted for finite samples, and accounts for seasonality.
- The Hodrick-Prescott filter depends on a parameter $q > 0$:
\[
 \psi (e^{-i \lambda}) = \frac{q}{q + {| 1 - e^{-i \lambda} |}^4 }.
\]
- There is a formula for the coefficients. We plot the filter coefficients with $q=1/1600$.

```{r}
q <- 1/1600

s <- (2*q + 2*q^(1/2)*(q+16)^(1/2))^(1/2)
r <- (q^(1/2) + (q+16)^(1/2) + s)/4
c <- q/r^2
phi1 <- 2*(q^(1/2)-(q+16)^(1/2))/(4*r)
phi2 <- (q^(1/2)+(q+16)^(1/2) - s)/(4*r)
theta <- atan(s/4)

lags <- seq(0,100)
psi <- 2*c*r^(4-lags)*sin(theta)*(r^2*sin(theta*(1+lags)) - sin(theta*(lags-1)))
psi <- psi/((1-2*r^2*cos(2*theta)+r^4)*(r^2-1)*(1-cos(2*theta)))
psi <- c(rev(psi),psi[-1])

plot(ts(psi,start=-101),xlab="Index",ylab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
```

- Next, we show the squared gain function.

```{r}
grid <- 1000
lambda <- pi*seq(0,grid+1)/grid
gain <- q/(q + (2 -2*cos(lambda))^2)
sq.gain <- gain^2

plot(ts(sq.gain,start=0,frequency=grid),xlab="Frequency",ylab="Squared Gain",
	yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
```
