---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 10-1: MA Identification

## Paradigm 10.1.2. Sequential Testing for the MA Order

- To identify an MA model, we must use the sample ACF rather than the true ACF
- Take as null hypothesis that the process is an MA(q), such that
\[
 H_0 :  \gamma (h) = 0 \quad \mbox{for all} \;  h > q.
\]
- We can use the estimate $\widehat{\gamma} (h)$ as test statistic, for any $h > q$.
- Under $H_0$, the asymptotic variance is $n^{-1}$ times
\[
   \tau^2_{\infty} =  \sum_{k = -\infty}^{\infty} { \gamma (k) }^2.
\]
- We utilize an estimator of $\tau^2_{\infty}$ based on concepts of the so-called *frequency domain*. We form the *periodogram* $I$ by computing
\[
  I(\lambda_{\ell}) = \sum_{k= -n+1}^{n-1} \widehat{\gamma} (k) \cos (\lambda_{\ell} k),
\]
 where $\lambda_{\ell} = 2 \pi \ell/n$ for some integer $-[n/2]-n+1 \leq \ell \leq [n/2]$.
- Then our estimator is
\[
 \widehat{\tau^2_{\infty}} = \frac{1}{2n} \, \sum_{\ell = -[n/2]-n+1}^{[n/2]}
  { I (\lambda_{\ell} )  }^2. 
\]
- Our normalized test statistic is 
\[
   \sqrt{n} \frac{ \widehat{\gamma} (h) }{ \widehat{\tau_{\infty} }},
\]
 which is asymptotically standard normal.
- Note that we have not addressed the multiple testing problem of multiple lags $h$!

## Example 10.1.3. Sequential Identification of Non-Defense Capitalization.
 
- We examine the Non-Defense Capitalization (New Orders) series.
- We take first differences because of non-stationarity.
- We plot the ACF and PACF.

```{r}
ndc <- read.table("Nondefcap.dat")
ndc <- ts(ndc[,2],start=c(1992,3),frequency=12,names= "NewOrders")
ndc.diff <- diff(ndc)
n <- length(ndc.diff)
acf(ndc.diff)
gamma.hat <- acf(ndc.diff,lag=n-1,type="covariance",plot=FALSE)$acf[,,1]
kappa.hat <- pacf(ndc.diff,lag=n-1)$acf[,,1]
```

- These plots indicate that an MA model may be appropriate, because the ACF seems to truncate whereas the PACF has geometric decay.
- We do sequential testing, examining lags 1 through 8. 
- Individual tests get rejected at lags 1 and 6. The lag 6 result is barely significant at the 5% level, and may be spurious due to multiple testing. 
- We tentatively conclude with an MA(1) model specification.

```{r}
lambda <- seq(-n/2+1,n/2)*2*pi/n
pgram <- cos(0*lambda)*gamma.hat[1]
for(h in 1:(n-1))
{
	pgram <- pgram + 2*cos(h*lambda)*gamma.hat[h+1]
}
pgram <- ts(pgram,start=0,frequency=n)
tau.hat <- sqrt(.5*sum(pgram^2)/n)
tstat <- sqrt(n)*gamma.hat[2:9]/tau.hat
round(rbind(tstat,rep(qnorm(1-.05/2),8)),digits=5)
```

## Paradigm 10.1.4. Joint Testing for the MA Order

- We want to test at multiple lags.
- $H_0$ is equivalent to $0 = \mbox{max} \{ | \rho (h) |, h > q \}$.
- **Empirical Rule**: let $\widehat{q}$ be the smallest positive integer such that 
\[
 |\widehat{\rho} (\widehat{q}+k) | < c \sqrt{n^{-1} \log n} 
\]
 for all $k = 1, \ldots, K_n$, where $c > 0$ is a fixed constant, and $K_n$ is a positive, non-decreasing integer-valued function of $n$ such that $K_n = o(\log n)$.
- $\widehat{q}$ is consistent for $q$.
- Choose $c = 1.96$ and $K_n = 1 + [3 \sqrt{ \log n}]$

## Example 10.1.5. Joint Testing Identification of Non-Defense Capitalization

- We now apply the Empirical Rule.
- $n =292$ implies $K_n = 5$.
- The first $5$ sample autocorrelations are given below.


```{r}
alpha <- .05
crit <- qnorm(1-alpha/2)
K.n <- 1 + floor(3*sqrt(log(n,base=10)))
rho.hat <- gamma.hat[-1]/gamma.hat[1]
print(rho.hat[1:K.n])
rho.test <- rho.hat/sqrt(log(n,base=10)/n)
k <- 1
while(k < (n-K.n))
{
	if(max(abs(rho.test[k:(k+K.n-1)])) < crit) 
	{ 
		q.hat <- k-1
		k <- n-K.n
	} else { k <- k+1 }
}
```

- The selected model order is `r q.hat`, which agrees with our results above.

