---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 9-1: Weak Dependence

- We can quantify serial dependence through the ACVF.
- Recall entropy mixing (Paradigm 8.2.11) as another measure of serial dependence.

## Example 9.1.1. Slow Polynomial Decay -- Long Memory

- Suppose that a time series has ACVF $\gamma (h) = O({|h|}^{-a})$ for $0 < a \leq 1$.
- The notation means the ACVF is bounded by a constant times that rate.
- There is fairly slow decay in lag when $a$ is small: this means persistence, or
high association, across large time lags. This is called *long-range dependence* or
*long memory*.
- For example, suppose $\gamma (h) = {|h|}^{-a}$ for $h \neq 0$ and 
$\gamma (0) = 2 \sum_{h =1}^{\infty} h^{-a}$ (which is twice the Riemann zeta function
at $a$). This value of $\gamma (0)$ ensures the ACVF is positive definite; however,
the ACVF is not summable, so the spectral density does not exist.

```{r}
#library(pracma)
lags <- seq(1,100)
a <- .4
gamma.lm <- c(2,lags^{-a})
plot(ts(gamma.lm,start=0),type="h",xlab="Lag",ylab="")
```

## Example 9.1.2. 
