---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---
 
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```
 
# Lesson 2-6: White Noise Processes

- *White noise* is a fundamental building block for time series models.
- Any $\{ X_t \}$ i.i.d. with mean zero and variance $\sigma^2$ is a $\mbox{WN}(0,\sigma^2)$.
- Here we provide three examples of white noise. 


## Example 2.6.1. Dependent White Noise
  
- Consider $X_t = Z_t \cdot Z_{t-1}$, where $Z_t$ is i.i.d. $N(0,1)$.  
- Then $X_t \sim \mbox{WN} (0,1)$, but $\{ X_t \}$ is not i.i.d.

```{r}
n <- 101
z <- rnorm(n)
x <- z[-n]*z[-1]
plot(ts(x),xlab="Time",ylab="")
#acf(x)
```

## Example 2.6.2. Non-identically Distributed White Noise

- Let $\{ Y_t \}$ and $\{ Z_t \}$ be independent of each other. Set $Z_t$ i.i.d. $N(0,1)$ and $Y_t$ i.i.d. uniform on $(-\sqrt{3},\sqrt{3})$. Let 
\[
 X_t = \begin{cases} Z_t \quad t \; \mbox{even} \\ Y_t \quad t \; \mbox{odd} \end{cases}
\]
- Then $X_t \sim \mbox{WN} (0,1)$, although the process is not stationary (since the marginal distribution depends on $t$).

```{r}
n <- 100
z <- rnorm(n)
y <- runif(n,-3^(1/2),3^(1/2))
x <- matrix(t(cbind(y,z)),ncol=1)
plot(ts(x),xlab="Time",ylab="")
#acf(x)
```


## Example 2.6.3. ARCH Process

- Model of Engel (1982), a Nobel laureate.
- Set $Z_t$ i.i.d. $N(0,1)$, and $X_t = Z_t \, \sqrt{\alpha + \beta X_{t-1}^2}$.
- Then $X_t \sim \mbox{WN} (0,\alpha/(1-\beta))$.

```{r}
n <- 101
z <- rnorm(n)
alpha <- .2
beta <- .3
x <- 0
for(t in 2:n) { x <- c(x,z[t]*sqrt(alpha + beta*x[t-1]^2)) }
plot(ts(x),xlab="Time",ylab="")
#acf(x)
```


