---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 7-6: Wold Decomposition

- We examine a decomposition of a stationary time series due to Wold.
- We also take a look at spectral factorization.

## Definition 7.6.1.

- A covariance stationary time series with innovation variance equal zero is called
**predictable**, since it can be forecasted without error.
- Any stochastic sinusoid is predictable: it is perfectly periodic!

## Theorem 7.6.4. Wold Decomposition.

A weakly stationary mean zero time series $\{ X_t \}$ that is not predictable can be
decomposed into a sum of processes $\{ U_t \}$ and $\{ Y_t \}$ such that:

- $X_t = U_t + Y_t$.
- $\{U_t \}$ and $\{ Y_t \}$ are uncorrelated with one another.
- $\{ U_t \}$ is predictable.
- $\{ Y_t \}$ has a MA($\infty$) representation $Y_t = \psi (B) Z_t$, 
with $Z_t \sim \mbox{WN} (0, \sigma^2)$.
- $\psi (z) = \sum_{j \geq 0} \psi_j z^j$ is causal with first coefficient $1$
($\psi (0) = 1$).
- $Z_t$ is the innovation at time $t$ of $\{ Y_t \}$ and $\{ X_t \}$
- $\sigma^2$ is the prediction error variance
- $\{ \psi_j \}$ are called the **Wold** coefficients

## Theorem D.2.7. Spectral Factorization

- Suppose $g(\lambda) = \sum_{k=0}^q g_k \cos (\lambda k)$ is a degree $q$ cosine polynomial
that is positive. Then there exists a degree $q$ polynomial $\theta (z)$ with unit
leading coefficient ($1 = \theta (0)$) and constant $\kappa >0$ such that
\[
 g(\lambda) = \kappa {| \theta (e^{-i \lambda}) |}^2.
\]
- This can be applied to the spectral density: suppose we know the autocovariances
$\gamma_k$ of an MA($q$) process, but not the MA coefficients. Then the spectral density is
a positive cosine polynomial of degree $q$, and we can find $\theta (z)$, the 
corresponding MA polynomial.
- For large $q$, these MA coefficients approximate the Wold coefficients.

## Exercise 7.44. Spectral Factorization of an MA(2) via Root-Finding.

- Use root-finding to do spectral factorization for an MA($2$) process.
- First write code:

```{r}
polymul <- function(a,b) 
{
	bb <- c(b,rep(0,length(a)-1))
	B <- toeplitz(bb)
	B[lower.tri(B)] <- 0
	aa <- rev(c(a,rep(0,length(b)-1)))
	prod <- B %*% matrix(aa,length(aa),1)
	return(rev(prod[,1]))
}
specfact.ma2 <- function(gamma)
{
	my.roots <-	polyroot(c(rev(gamma),gamma[-1]))
	ind.inv <- which(Mod(my.roots)>1)	
	theta.pol <- polymul(c(1,-1/my.roots[ind.inv[1]]),c(1,-1/my.roots[ind.inv[2]]))	
	theta.pol <- Re(theta.pol)
	kappa <- Re(gamma[3]*my.roots[ind.inv[1]]*my.roots[ind.inv[2]])
	return(list(theta.pol,kappa))
}
```

- Test this out on an MA(2) with $\gamma_0 = 1$, $\gamma_1 = .6$, and $\gamma_2 = .2$. 

```{r}
gamma <- c(1,.6,.2)
out <- specfact.ma2(gamma)
theta.pol <- out[[1]]
kappa <- out[[2]]
print(c(theta.pol))
print(kappa)
## check
print(polymul(theta.pol,rev(theta.pol))*kappa)
```



