---
title: 'Time Series: A First Course with Bootstrap Sampler'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/TimeSeriesR/Data')
```

# Lesson 10: Seasonality

- Regular patterns of a periodic nature are called *seasonality*. 
- The same notion as a *cycle*, but is associated with the calendar.
- We may wish to estimate and remove seasonality in order to see other patterns.

## Example 3.5.1. Mauna Loa Growth Rate

- We consider the Mauna Loa CO2 time series.

```{r}
mau <- read.table("mauna.dat",header=TRUE,sep="")
mau <- ts(mau,start=1958,frequency=12)
plot(mau,xlab="Year",ylab="Mauna Loa")
```

- We can difference the series to eliminate the trend.
- A pronounced seasonal pattern is apparent; but it is not completely periodic.

```{r}
mau.diff <- diff(mau)
plot(mau.diff,xlab="Year",ylab="Mauna Loa")
```

## Seasonal Regression

- An exactly periodic effect would be $X_t = X_{t-s}$ for seasonal period $s$ (e.g., $s=12$ for monthly time series), and any $t$.
- If this were true, we could use dummy regressors. Usually this works poorly; if the series were periodic, then the filter $1 - B^s$ would annihilate it.
- Example of Mauna Loa; seasonal differencing of the growth rate leaves a non-zero time series.

```{r}
mau.diffs <- diff(mau.diff,lag=12)
plot(mau.diffs,xlab="Year",ylab="Mauna Loa")
```

## Seasonal Moving Average

- A better way is to estimate seasonality using a *seasonal moving average*. This is a moving average filter where only coefficients with indices that are a multiple of $s$ are non-zero.
- This means 
\[
   \Psi (B) = \sum_{k = -ms}^{ms} \psi_k B^k
    = \sum_{j=-m}^m \psi_{js} B^{js}
\]
 is a seasonal moving average, because $\psi_k = 0$ unless $k = js$ for some integer $j$.
 
## Example 3.5.8. Seasonal Moving Average for Motor Retail Sales

- We apply a seasonal moving average to the growth rate of the Motor Retail Sales data, with $m=3$.
- We plot the growth rate series, the estimated seasonality, and the residual after subtracting off the seasonal component. 

```{r}
Ret441 <- read.table("retail441.b1",header=TRUE,skip=2)[,2]
Ret441 <- ts(Ret441,start = 1992,frequency=12)
ret.diff <- diff(log(Ret441))
 
pseas <- 3
period <- 12
seas.smoother <- c(1,rep(0,period-1))
seas.smoother <- c(rep(seas.smoother,pseas),rep(seas.smoother,pseas),1)/(2*pseas+1)
seas.smooth <- filter(ret.diff,seas.smoother,method="convolution")
seas.resid <- ret.diff - seas.smooth

par(oma=c(2,0,0,0),mar=c(2,4,2,2)+0.1,mfrow=c(3,1),cex.lab=1.2)
plot(ret.diff,ylab="Motor Growth Rate",xlab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(seas.smooth,start=1992,frequency=12),xlab="",ylab="Seasonal",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(seas.resid,start=1992,frequency=12),xlab="",ylab="Residual",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
mtext(text="Year",side=1,line=1,outer=TRUE)
```


## Remark 3.5.9. Seasonal Adjustment

- Removal of seasonality is called *seasonal adjustment*.
- The *seasonal aggregation* filter is defined as
\[
  U(B) = 1 + B +  B^2 + \ldots + B^{s-1}.
\]
 It annihilates any periodic pattern, and is a basic example of a seasonal adjustment filter.
 
## Example 3.5.10. Seasonal Adjustment of Mauna Loa Growth Rate

- We apply the seasonal aggregation filter to the growth rate of the Mauna Loa CO2 time series.
- This uses a trick: applying $U(B)$ and $1-B$ together amounts to applying $1-B^s$, because
\[
  U(B) (1 - B) = 1- B^s.
\]

```{r}
seas.elim <- diff(mau,lag=12)
plot(seas.elim,xlab="Year",ylab="Mauna Loa")
```

## Paradigm 3.6.8. Classical Decomposition from Trend Filtering

- Consider a time series with both trend and seasonal effects.
- We can design a trend filter that eliminates seasonality, obtaining $\widehat{T}_t$.
- Then apply a seasonal filter to $X_t - \widehat{T}_t$, the de-trended data, obtaining $\widehat{S}_t$.
- Removing $\widehat{S}_t$ from the de-trended data yields the irregular $\widehat{I}_t$.
- Then we have the *classical decomposition*
\[
  X_t = \widehat{T}_t + \widehat{S}_t + \widehat{I}_t.
\]

## Example 3.6.13. Classical Decomposition of Western Housing Starts via Linear Filtering

- We determine a classical decomposition by using a trend filter 
\[
   \frac{1}{2s} U(B) (1+B) B^{-s/2}
   = \frac{1}{2s} B^{-s/2} + \frac{1}{s} B^{-s/2+1} + \ldots + \frac{1}{s} B^{s/2-1} + 
     \frac{1}{2s} B^{s/2},
\]
 when $s$ is even.  
 
```{r}
compSmooth <- function(data,period,pseas)
{
	# compSmooth
	#	Estimates trend, seasonal, and irregular from input data
	#	using weighted moving averages.
	#	data: an input time series
	#	period: observations per year
	#	pseas: odd integer giving number of years in seasonal moving average
	#	

	n <- length(data)
	trendfilter <- seq(1,period)
	trendfilter <- c(trendfilter[1:(period-1)],rev(trendfilter))/period^2
	seasfilter <- c(1,rep(0,period-1))
	seasfilter <- c(rep(seasfilter,pseas),rep(seasfilter,pseas),1)/(2*pseas+1)
	trend <- filter(data,trendfilter,method="convolution",sides=2)
	trend <- trend[period:(n-period+1)]
	resid <- data[period:(n-period+1)] - trend
	seasonal <- filter(resid,seasfilter,method="convolution",sides=2)
	seasonal <- seasonal[(pseas*period + 1):(length(resid) - pseas*period)]
	irreg <- resid[(pseas*period + 1):(length(resid) - pseas*period)] - seasonal

	return(cbind(trend[(pseas*period + 1):(length(resid) - pseas*period)],seasonal,irreg))
}

Wstarts <- read.table("Wstarts.b1",skip=2)[,2]
Wstarts <- ts(Wstarts,start = 1964,frequency=12)
pseas <- 3
comps <- compSmooth(Wstarts,12,pseas)

par(oma=c(2,0,0,0),mar=c(2,4,2,2)+0.1,mfrow=c(4,1),cex.lab=1.2)
plot(Wstarts,ylab="Starts",xlab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(c(rep(NA,47),comps[,1],rep(NA,47)),start=1964,frequency=12),xlab="",ylab="Trend",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(c(rep(NA,47),comps[,2],rep(NA,47)),start=1964,frequency=12),xlab="",ylab="Seasonal",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(c(rep(NA,47),comps[,3],rep(NA,47)),start=1964,frequency=12),xlab="",ylab="Irregular",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
mtext(text="Year",side=1,line=1,outer=TRUE)
```
 
 