---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 8-6: Modeling via Entropy

- We can attempt to model data so as to increase entropy.

## Definition 8.6.2.

Given a sample $\underline{X}$ with pdf $p_{\underline{X}}$, a **model** is the 
composition $\Pi$ of successive entropy-increasing transformations, such that the
**residuals** $\underline{Z} = \Pi [ \underline{X} ]$ has maximum entropy among
the class of transformations. 

## Example 8.6.7. Log Difference

- The growth rate transformation (log differencing) can sometimes increase entropy
for economic time series.
- Consider $X_t = \exp Z_t$ where $\{ Z_t \}$ is a random walk. Log differencing is
a model $\Pi$ that maps the process to white noise:
\[
  \log X_t - \log X_{t-1} = Z_t - Z_{t-1} = \epsilon_t.
\]

## Example 8.6.8. Entropy-Increasing Transformation for U.S. Population

- The raw data of U.S. population has a lot of structure (low entropy).
- We know that first differences remove much of the trend structure. So $1-B$ is
a model for the data.
- We can also consider the model ${(1-B)}^2$.

```{r}
pop <- read.table("USpop.dat")
pop <- ts(pop, start = 1901)
diff.pop <- diff(pop*10e-6)
diffdiff.pop <- diff(diff(pop*10e-6))
plot(diff.pop,xlab="Year",ylab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(diffdiff.pop,xlab="Year",ylab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
```

## Exercise 8.40. Entropy-Increasing Transformation of Electronics and Appliance Stores

- What is an entropy-increasing transformation (or model) for the dataset of Electronics
and Appliance Stores?
- First plot the data.

```{r}
ret <- read.table("retail443.b1",header=FALSE,skip=2)[,2]
ret <- ts(ret,start=1992,frequency=12)
plot(ret)
```

- Examine the log transformation.

```{r}
ret.ent <- ts(log(ret),start=start(ret),frequency=12)
plot(ret.ent)
```

- Examine log differences.

```{r}
ret.ent <- ts(diff(log(ret)),start=start(ret),frequency=12)
plot(ret.ent)
```

- Examine logs with seasonal differencing.

```{r}
ret.ent <- ts(diff(log(ret),lag=12),start=start(ret),frequency=12)
plot(ret.ent)
```

- Examine logs with seasonal and nonseasonal differencing.

```{r}
ret.ent <- ts(diff(diff(log(ret),lag=12)),start=start(ret),frequency=12)
plot(ret.ent)
```

