---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 4-1: Vector Space Geometry

- Euclidean geometry and linear algebra are tools for analyzing n-dimensional space.
- We can adapt these tools to studying time series random vectors. 

## Example 4.1.1.  Angle Between Two Vectors

- Let $\underline{x} = [1, \sqrt{3}]$ and $\underline{y} = [1,1]$
- Let $\theta$ be the angle between them.
- Their angles with the x-axis are $\pi/3$ and $\pi/4$ respectively. So $\theta = \pi/12$.

```{r}
x <- c(1,1)
y <- c(1,sqrt(3))
  
par(mar=c(4,4,2,2)+0.1,cex.lab=.8)
plot(NA,xlim=c(-.2,2),ylim=c(-.2,2),xlab="x-axis",ylab="y-axis",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
x0 <- c(0,0,1,sqrt(3))
y0 <- c(0,0,1,1)
arrows(x0[1],x0[2],x0[3],x0[4],col=1,lwd=2)
arrows(y0[1],y0[2],y0[3],y0[4],col=1,lwd=2)
text(1.1,sqrt(3)-.05,"x",cex=1.5)
text(1.1,.95,"y",cex=1.5)
```

## Inner Product

- Measure a degree of similarity of two vectors via the *inner product*.
- For vectors $\underline{x}, \underline{y} \in {\mathbb R}^n$, their inner product is
\[
 \langle \underline{x}, \underline{y} \rangle = \sum_{i=1}^n x_i y_i.
\]
- Also, $\| \underline{x} \| = \sqrt{ \langle \underline{x}, \underline{x} \rangle }$ is the norm of $\underline{x}$.
- The angle $\theta$ between these two vectors satisfies
\[
 \cos (\theta) = \frac{\langle \underline{x}, \underline{y} \rangle }{ \| \underline{x} \| \| \underline{y} \|}.
\]

## Theorem 4.1.7. Cauchy-Schwarz Inequality

- For $\underline{x}, \underline{y}$ in a vector space with inner product,
\[
 | \langle \underline{x}, \underline{y} \rangle | \leq \| \underline{x} \| \| \underline{y} \|.
\]
- Equality occurs if and only if the vectors are a scalar multiple of one another.



