---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 7-7: Cepstrum

- The AR and MA (and ARMA) processes are dense in the class of covariance stationary
processes, in the sense that an arbitrary continuous spectral density can be 
approximated by an MA($q$), AR($p$), or ARMA($p$,$q$) spectral density for sufficiently
large $p$ and $q$.
- Another class of spectral density is given by the "cepstrum".

## Paradigm 7.7.5. The Cepstrum

- We find the Fourier transform of the log of a positive spectral density:
\[
\log f(\lambda) = \sum_{k = - \infty}^{\infty} \tau_k e^{-i \lambda k}.
\]
- The Fourier coefficients $\{ \tau_k \}$ are called **cepstral** coefficients.
- This representation of $\log f$ is called the **cepstrum**; the word "cepstrum" 
is an anagram of "spectrum".
- $\log f (\lambda)$ can be negative.
- The cepstral coefficients are recovered via Fourier inversion:
\[
 \tau_k = \frac{1}{ 2 \pi} \int_{-\pi}^{\pi} e^{i \lambda k} \log f(\lambda) \, d\lambda.
\]
- Since $f$ is even, $\tau_{-k} = \tau_k$.
- Kolmogorov: $\tau_0 = \log \sigma^2$.

## Proposition 7.7.7. Cepstral Representation of Wold Coefficients

- We can relate the cepstral coefficients to the Wold coefficients, since
\[
 f(\lambda) = \exp \left\{ \sum_{k = - \infty}^{\infty} \tau_k e^{-i \lambda k} \right\}
  = e^{\tau_0} \cdot \exp \left\{ \sum_{k > 0} \tau_k e^{-i \lambda k} \right\}
   \cdot \exp \left\{ \sum_{k > 0} \tau_k e^{i \lambda k} \right\},
\]
which suggests
\[
\exp \left\{ \sum_{k > 0} \tau_k z^k \right\} = \sum_{j \geq 0} \psi_j z^j,
\]
where $\{ \psi_j \}$ are Wold coefficients.
- This relationship provides a recursive formula for Wold in terms of cepstral:
\[
 \psi_j = \frac{1}{j} \sum_{k=1}^j k \, \tau_k \, \psi_{j-k}
\]
for $j \geq 1$, which is initialized with $\psi_0 = 1$. 

## Definition 7.7.9.

- The **exponential process** or order $m$, or EXP($m$), is a causal time series $\{ X_t\}$
with spectral density $f$ given by
\[
  f(\lambda) = \exp \left\{ \sum_{|k| \leq m} \tau_k e^{-i \lambda k} \right\}.
\]
- All EXP processes are invertible!
- This defines a class of models, just like AR and MA models.
- The EXP class is also dense.

## Exercise 7.46. Moving Average Representation of Cepstral Process

- We encode the cepstral to Wold mapping, and apply to an EXP(2) process.
- Our function takes a sequence $\tau_1, \ldots, \tau_m$ as inputs, and generates
$\psi_1, \ldots, \psi_q$ for given $q$.

```{r}
ceps2wold <- function(ceps,q)
{
	m <- length(ceps)
	if(q > m) {	ceps <- c(ceps,rep(0,q-m)) }
	wold <- 1
	wolds <- wold
	for(j in 1:q)
	{
		wold <- sum(seq(1,j)*ceps[1:j]*wolds[j:1])/j
		wolds <- c(wolds,wold)
	}
	return(wolds)
}
```

- We apply this with $\tau_1 = .5$ and $\tau_2 = -.2$.

```{r}
print(round(ceps2wold(c(.5,-.2),20),digits=4))
```

## Exercise 7.52. Spectral Factorization of an MA($q$) via Cepstrum

- We can use the cepstrum to do a spectral factorization, as an alternative method
to the root-finding approach of Exercise 7.44.
- We take as inputs the autocovariances, numerically compute the cepstral coefficients,
and then apply Proposition 7.7.7 to get the Wold coefficients.

```{r}
specceps.maq <- function(gamma,mesh)
{
	lambda <- pi*seq(0,mesh)/mesh
	q <- length(gamma)-1
	f.spec <- gamma[1]*cos(0*lambda) 
	for(h in 1:q)
	{
		f.spec <- f.spec + 2*gamma[h+1]*cos(h*lambda)
	}
	theta.pol <- 1
	kappa <- 1
	if(min(f.spec)>0)
	{
		tau <- mean(log(f.spec))
		taus <- tau
		for(h in 1:q)
		{
			tau <- mean(log(f.spec)*cos(h*lambda))
			taus <- c(taus,tau)
		}
		theta.pol <- ceps2wold(taus[-1],q)
		kappa <- exp(taus[1])
	}
 	return(list(theta.pol,kappa))
}
```

- This function computes cepstral coefficients. 
- We test on the MA(5) with $\theta (z) = 1 + .8z + .6 z^2 + .3z^3 + .5z^4 + .2z^5$
with $\sigma^2 = 8$, and show that the MA coefficients are recovered.

```{r}
polymul <- function(a,b) 
{
	bb <- c(b,rep(0,length(a)-1))
	B <- toeplitz(bb)
	B[lower.tri(B)] <- 0
	aa <- rev(c(a,rep(0,length(b)-1)))
	prod <- B %*% matrix(aa,length(aa),1)
	return(rev(prod[,1]))
}
theta.pol <- c(1,.8,.6,.3,.5,.2)
sig2 <- 8
gamma <- polymul(theta.pol,rev(theta.pol))*sig2
gamma <- gamma[6:11]
print(gamma)
out <- specceps.maq(gamma,10000)
theta.pol <- out[[1]]
kappa <- out[[2]]
print(theta.pol)
print(kappa)
```

- We can repeat with a larger mesh size, to get a closer approximation.


