---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 7-2: Discrete Fourier Transform

We introduce the data analysis tool of Discrete Fourier Transform.

## Definition 7.2.3.

Given a sample $X_1, \ldots, X_n$, the **Discrete Fourier Transform** (DFT) at the
Fourier frequency $\lambda_l = 2 \pi l/n$ (for $[n/2]-n+1 \leq l \leq [n/2]$) is
\[
 \widetilde{X} (\lambda_l) = n^{-1/2} \sum_{t=1}^n X_t e^{-i \lambda_l t}.
\]

## Definition 7.2.4.

- The **periodogram** $I(\lambda)$ is a non-negative function of $\lambda \in [-\pi,\pi]$,
constructed from the sample $X_1, \ldots, X_n$:
\[
 I(\lambda) = n^{-1} { \left| \sum_{t=1}^n (X_t - \overline{X}) e^{-i \lambda t} \right| }^2.
\]
- So $I(0) = 0$. Sometimes we consider an "uncentered" periodogram, where there is
no centering by the sample mean. We denote this by $\widetilde{I} (\lambda)$, and
\[
  \widetilde{I} (\lambda_l) = {| \widetilde{X} (\lambda_l) |}^2.
\]
- The periodogram is an empirical version of the spectral density, and shares certain
properties. Higher values correspond to cyclical effects in the data.

## Proposition 7.2.7.

- Let the vector of DFTs be denoted $\widetilde{\underline{X}}$, which has components
$\widetilde{X} (\lambda_l)$. 
- This is a linear function of the sample vector $\underline{X}$: 
\[
\widetilde{\underline{X}} =  Q^* \, \underline{X},
\]
where $Q$ was defined in Definition 6.4.3. 
- Since $Q$ is unitary, $Q^{-1} = Q^*$, so we can recover the data from the DFT vector
via
\[
  \underline{X} = Q \, \widetilde{\underline{X}}.
\]

## Exercise 7.14. DFT of an AR(1).

- We simulate an AR(1).

```{r}
arp.sim <- function(n,burn,ar.coefs,innovar)
{
	p <- length(ar.coefs)
	z <- rnorm(n+burn+p,sd=sqrt(innovar))
	x <- z[1:p]
	for(t in (p+1):(p+n+burn))
	{
		next.x <- sum(ar.coefs*x[(t-1):(t-p)]) + z[t]
		x <- c(x,next.x)
	}	
	x <- x[(p+burn+1):(p+burn+n)]
	return(x)
}

phi <- .8
innovar <- 1
n <- 200
x.sim <- arp.sim(n,500,phi,innovar)
```

- We compute the DFT. We begin with code to compute $Q$.

```{r}

get.qmat <- function(mesh)
{
	mesh2 <- floor(mesh/2)
	inds <- seq(mesh2-mesh+1,mesh2)
	Q.mat <- exp(1i*2*pi*mesh^{-1}*t(t(seq(1,mesh)) %x% inds))*mesh^{-1/2}
	return(Q.mat)
}
Q.mat <- get.qmat(n)
```

- Then we use Proposition 7.2.7 to get the DFT vector. We plot the real part, imaginary part,
and the modulus.

```{r}
x.dft <- Conj(t(Q.mat)) %*% x.sim
#par(mfrow=c(3,1))
plot(ts(Re(x.dft),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Real Part")
plot(ts(Im(x.dft),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Imaginary Part")
plot(ts(Mod(x.dft)^2,start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Squared Modulus")
```

## Corollary 7.2.9. Decorrelation Property of the DFT.

- Let $X_1, \ldots, X_n$ be a sample from a mean zero, covariance stationary time
series with absolutely summable ACVF and spectral density $f$. Then 
$\widetilde{\underline{X}}$ has approximate covariance matrix 
$\mbox{diag} \{  f(\lambda_{[n/2]-n+k}) \}$.
- This follows from Theorem 6.4.5:
\[
 \mbox{Cov} [ \widetilde{\underline{X}} ] = Q^* \, 
 \mbox{Cov} [ \underline{X} ] \, Q  \approx \Lambda.
\]

## Exercise 7.18. DFT of Wolfer Sunspots.

- We compute the DFT of the Wolfer sunspot data, and plot.

```{r}
wolfer <- read.table("wolfer.dat")
wolfer <- ts(wolfer,start=1749,frequency=12)
n <- length(wolfer)
Q.mat <- get.qmat(n)
x.dft <- Conj(t(Q.mat)) %*% wolfer
plot(ts(Re(x.dft),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Real Part")
plot(ts(Im(x.dft),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Imaginary Part")
plot(ts(log(Mod(x.dft)^2),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Log Squared Modulus")
```

- We see higher values of the squared modulus (the uncentered periodogram) near
frequency zero.

## Exercise 7.20. DFT of Mauna Loa Growth Rate.

- We compute the DFT of the Mauna Loa growth rate, and plot.

```{r}
mau <- read.table("mauna.dat",header=TRUE,sep="")
mau <- ts(mau,start=1958,frequency=12)
mau.gr <- diff(log(mau))
n <- length(mau.gr)
Q.mat <- get.qmat(n)
x.dft <- Conj(t(Q.mat)) %*% mau.gr
plot(ts(Re(x.dft),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Real Part")
plot(ts(Im(x.dft),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Imaginary Part")
plot(ts(log(Mod(x.dft)^2),start=-floor(n/2)+1,frequency=1),
	xlab="Frequency Index",ylab="Log Squared Modulus")
```

- We see higher values of the uncentered periodogram in the shape of peaks, at four
non-zero frequencies. These correspond to cyclical seasonal effects.
