---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 3-6: Trend and Seasonality Together

- We may want to extract both trend and seasonality from time series.
- This is more challenging than just extracting trend or seasonality alone.

## Proposition 3.6.5

- Stable seasonality has an exactly periodic pattern.
- A trend extraction filter $\Psi (B)$ eliminates stable seasonality if it can be
written as
\[
 \Psi (B) = \Theta (B) U(B),
\]
 for some filter $\Theta (B)$, where $U(B)$ is the seasonal aggregation filter.
 
## Example 3.6.7. Trend Filters that Eliminate Seasonality

- For $s$ even, consider the trend filter 
\[
\Psi (B) = \frac{1}{2s} B^{-s/2} + \frac{1}{s} B^{-s/2+1} + \ldots + \frac{1}{s} B^{s/2-1} + 
     \frac{1}{2s} B^{s/2}.
\]
- By polynomial multiplication, we can show that
\[
 \Psi (B) = \frac{1}{2s} U(B) (1+B) B^{-s/2}.
\]
- Setting $\Theta(B) = (1+B) B^{-s/2}/(2s)$, we see that $\Psi (B)$ eliminates stable seasonality.

## Paradigm 3.6.8. Classical Decomposition from Trend Filtering

- Consider a time series with both trend and seasonal effects.
- We can design a trend filter that eliminates seasonality, obtaining $\widehat{T}_t$.
- Then apply a seasonal filter to $X_t - \widehat{T}_t$, the de-trended data, obtaining $\widehat{S}_t$.
- Removing $\widehat{S}_t$ from the de-trended data yields the irregular $\widehat{I}_t$.
- Then we have the *classical decomposition*
\[
  X_t = \widehat{T}_t + \widehat{S}_t + \widehat{I}_t.
\]

## Example 3.6.13. Classical Decomposition of Western Housing Starts via Linear Filtering

- Consider the trend filter given in Example 3.6.7.
- We construct the classical decomposition, applied to Western Housing Starts
 
```{r}
compSmooth <- function(data,period,pseas)
{
	# compSmooth
	#	Estimates trend, seasonal, and irregular from input data
	#	using weighted moving averages.
	#	data: an input time series
	#	period: observations per year
	#	pseas: odd integer giving number of years in seasonal moving average
	#	

	n <- length(data)
	trendfilter <- seq(1,period)
	trendfilter <- c(trendfilter[1:(period-1)],rev(trendfilter))/period^2
	seasfilter <- c(1,rep(0,period-1))
	seasfilter <- c(rep(seasfilter,pseas),rep(seasfilter,pseas),1)/(2*pseas+1)
	trend <- filter(data,trendfilter,method="convolution",sides=2)
	trend <- trend[period:(n-period+1)]
	resid <- data[period:(n-period+1)] - trend
	seasonal <- filter(resid,seasfilter,method="convolution",sides=2)
	seasonal <- seasonal[(pseas*period + 1):(length(resid) - pseas*period)]
	irreg <- resid[(pseas*period + 1):(length(resid) - pseas*period)] - seasonal

	return(cbind(trend[(pseas*period + 1):(length(resid) - pseas*period)],seasonal,irreg))
}

Wstarts <- read.table("Wstarts.b1",skip=2)[,2]
Wstarts <- ts(Wstarts,start = 1964,frequency=12)
pseas <- 3
comps <- compSmooth(Wstarts,12,pseas)

par(oma=c(2,0,0,0),mar=c(2,4,2,2)+0.1,mfrow=c(4,1),cex.lab=1.2)
plot(Wstarts,ylab="Starts",xlab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(c(rep(NA,47),comps[,1],rep(NA,47)),start=1964,frequency=12),xlab="",ylab="Trend",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(c(rep(NA,47),comps[,2],rep(NA,47)),start=1964,frequency=12),xlab="",ylab="Seasonal",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
plot(ts(c(rep(NA,47),comps[,3],rep(NA,47)),start=1964,frequency=12),xlab="",ylab="Irregular",yaxt="n",xaxt="n")
axis(1,cex.axis=.8)
axis(2,cex.axis=.8)
mtext(text="Year",side=1,line=1,outer=TRUE)
```
 
 
