---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/home/tucker/Documents/GitHub/BEwebinar/Data')
```

# Lesson 9-5: Sample Autocovariance

- We renormalize the ACVF estimator via tapering.

## Remark 9.5.2. Tapering the ACVF to Reduce Variability in High Lags

- The estimator $\widetilde{\gamma} (h)$ is approximately unbiased, but has high variance
for large $h$.
- We can replace the $n-h$ divisor by $n$, in order to decrease variance for larger $h$.
- This is like multiplying $\widetilde{\gamma} (h)$ by $1 - h/n$, which is an example
of a taper.

## Definition 9.5.3.

- An autocovariance *taper* is a bounded, even real-valued function $\Lambda$ on $[-1,1]$,
such that $\Lambda (0) = 1$ and $\Lambda (x) \leq 1$. 
- We multiply the ACVF estimator at lag $h$ by $\Lambda (h/n)$, where $\Lambda (x) = 1 - |x|$.

## Example 9.5.4. Bartlett Taper

- The Bartlett taper is $\Lambda (x) = 1 - |x|$. 
- The resulting estimator is
\[
 \widehat{\gamma} (h) = \Lambda (h/n) \widetilde{\gamma} (h) = 
 \frac{1}{n} \sum_{t=1}^{n-|h|} (X_{t+|h|} - \overline{X})(X_t - \overline{X}).
\]
- This is called the *sample autocovariance function* (sample ACVF) 
- Under $m$-dependence, Corollary 9.5.8 establishes a central limit theorem for the sample ACVF:
  \[
   \sqrt{n} ( \widehat{\gamma} (h)  - \gamma (h)) \Rightarrow \mathcal{N} (0, \tau^2_{\infty}).
\]

## Exercise 9.28. Sample ACVF of ARMA(1,2) Process

- We simulate a Gaussian ARMA(1,2) process of length $n=200$.

```{r}
armapq.sim <- function(n,burn,ar.coefs,ma.coefs,innovar)
{
	p <- length(ar.coefs)
	q <- length(ma.coefs)
	z <- rnorm(n+burn+p+q,sd=sqrt(innovar))
	x <- filter(z,c(1,ma.coefs),method="convolution",sides=1)
	x <- x[(q+1):(q+n+burn+p)]
	y <- x[1:p]
	for(t in (p+1):(p+n+burn))
	{
		next.y <- sum(ar.coefs*y[(t-1):(t-p)]) + x[t]
		y <- c(y,next.y)
	}	
	y <- y[(p+burn+1):(p+burn+n)]
	return(y)
}

n <- 200
phi1 <- .5
theta1 <- 5/6
theta2 <- 1/6
x.sim <- armapq.sim(n,500,phi1,c(theta1,theta2),1)
```

- We construct and plot the sample acvf.

```{r}
y.sim <- x.sim - mean(x.sim)
x.acf <- mean(y.sim^2)
for(h in 1:20)
{
	x.acf <- c(x.acf,sum(y.sim[1:(n-h)]*y.sim[(h+1):n])/n)
}
plot(ts(x.acf,start=0,frequency=1),xlab="Lag",ylab="Sample Autocovariance",type="h")
```
