---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/home/tucker/Documents/GitHub/BEwebinar/Data')
```

# Lesson 9-9: Spectral Analysis

- We refine the analysis of spectral density estimation.
- We consider the class of flat-top tapers.

## Paradigm 9.9.1. An Interesting Class of Tapers

- Consider a taper that takes the value $1$ in a neighborhood of zero: this improves the bias.
- This suggests the definition $\Lambda_{\infty} (x) = 1$ for $|x| \leq c$, for some $c \in (0,1]$. We call these *flat-top* tapers.
- Example: Bartlett taper (Example 9.5.4) is a limiting case with $c=0$.
- Example: rectangular taper with $c=1$, where $\Lambda (x) = 1$ for $|x| \leq 1$, zero otherwise.
- Example: trapezoidal taper (depending on choice of $c$):
\[
 \Lambda_{\infty} (x) = \begin{cases} 1 \quad \mbox{if} \; |x| \leq c \\
  \frac{1 - |x|}{1 - c} \quad \mbox{if} \; c < |x| \leq 1.  \end{cases}
\]
- Flat-top tapers correct bias by shifting the spectral estimate down, which can generate negative estimates. One can take the maximum with zero, so as to enforce a non-negative spectrum estimate.

## Example 9.9.6. Application of a Trapezoidal Taper

- We apply a trapezoidal taper to the Wolfer sunspots, using $c=1/3$ and $d = 42$.
- The spectral estimate has more variability, and is no longer positive (log of negative values is cut off).

```{r}
wolfer <- read.table("wolfer.dat")
wolfer <- ts(wolfer[,1],start=1749,frequency=12)
data <- wolfer
n <- length(data)
d <- 3*floor(n^{1/3})
acfs.sample <- NULL
mu.hat <- sum(data)/n
for(k in 0:(n-1))
{
	acf.sample <- sum((data[1:(n-k)]-mu.hat)*(data[(k+1):n]-mu.hat))/n
	acfs.sample <- c(acfs.sample,acf.sample)
}
cutoff <- 1/3
trap <- function(h)
{
	val <- min(max((1 - h/d)/(1 - cutoff),0),1)
	return(val)
}
grid <- 10000
lambda <- seq(0,grid)*pi/grid
pgram <- cos(0*lambda)*acfs.sample[1]
for(h in 1:(n-1))
{
	pgram <- pgram + 2*trap(h)*cos(h*lambda)*acfs.sample[h+1]
}
pgram <- pmax(0,pgram)
pgram <- ts(pgram[-1],start=0,frequency=grid)
plot(log(pgram),xlab="Cycles",ylab="Log Spectrum",ylim=c(3,12))
```
