---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 5-2: Difference Equations

- To understand ARMA processes it is useful to study *difference* equations, 
which are a discrete analogue of differential equations.

## Definition 5.2.2. 

- The equation $\phi (B) X_t = W_t$ is a *linear ordinary difference equation* (ODE)
for $\{ X_t \}$ with input $\{ W_t \}$.
- If $\phi (z)$ has degree $p$, then the ODE has order $p$.
- If $W_t \equiv 0$, the ODE is *homogeneous*.

## Paradigm 5.2.7. Solution of a Homogeneous ODE.

- The key is to obtain the roots of $\phi (z)$. 
- Why: let $\zeta$ be a root, so that $\phi (\zeta) = 0$. Then check that $X_t = \zeta^{-t}$
solves the homogeneous ODE:
\[
 \phi (B) X_t = \sum_{j=0}^p \phi_j X_{t-j} = \sum_{j=0}^p \phi_j \zeta^{-t+j}
 = \zeta^{-t} \, \sum_{j=0}^p \phi_j \zeta^j = \zeta^{-t} \, \phi (\zeta) = 0.
\]
- The polynomial has $p$ roots $\zeta_1, \ldots, \zeta_p$. These can be complex
numbers, and might be repeated or distinct.
- If the roots are distinct, the general solution has the format
\[
X_t = \sum_{j=1}^p b_j \zeta_j^{-t},
\]
 where $b_j$ are coefficients to be determined by the initial conditions.
- Initial conditions are specified values for $X_1, \ldots, X_p$. If these are real,
then the solution $X_t$ for $t > p$ will also be real. (Though the coefficients
$b_j$ might be complex.)

## Example 5.2.10. Fibonacci Sequence.

- Consider the Fibonacci recursion $X_t = X_{t-1} + X_{t-2}$, which corresponds to
an ODE with $\phi (z) = 1 - z - z^2$.
- The roots are distinct and real: $\zeta_1 = -1/2 + \sqrt{5}/2$, $\zeta_2 = -1/2 - \sqrt{5}/2$.
- With initial conditions $X_1 = X_0 = 1$, the coefficients are found to be 
$b_1 = -\zeta_2/\sqrt{5}$ and $b_2 = \zeta_1/\sqrt{5}$.
- Then $X_t = b_1 \zeta_1^{-t} + b_2 \zeta_2^{-t}$ is the homogeneous solution.
- We plot the sequence (in log scale) with this initialization.

```{r}
n <- 10
x <- rep(1,n)
for(i in 3:n) { x[i] <- x[i-1] + x[i-2] }
zeta1 <- (-1 + 5^{1/2})/2
zeta2 <- (-1 - 5^{1/2})/2
b1 <- -zeta2/5^{1/2}
b2 <- zeta1/5^{1/2}
y <- b1*zeta1^{-seq(0,n-1)} + b2*zeta2^{-seq(0,n-1)}
plot(ts(log(x)),xlab="Index",ylab="log Fibonnaci")
lines(ts(log(y)),col=2)
```


## Example 5.2.11. Seasonal Difference.

- Any periodic function can be written as a sum of cosines and sines. Why?
- If $X_t$ is periodic with integer period $s$, then it is annihilated by seasonal 
differencing, and $(1-B^s) X_t = 0$.
- So $\phi(z) = 1 - z^s$, which has roots $\zeta_j = e^{2 \pi i j/s}$ for $j=1,\ldots,s$.
- Thus the solution is
\[
 X_t = \sum_{j=1}^s b_j e^{-2 \pi i j t/s}.
\]
- If $s$ is even, then two roots (corresponding to $j = s/2, s$) are real, and the rest
are complex conjugate pairs. 
- If $s$ is odd, then one root (corresponding to $j=s$) is real, and the rest are
complex conjugate pairs.
- The sequence $X_t$ is real, so $b_j e^{-2 \pi i j t/s}$ must be real, and hence
\[
X_t = \sum_{j=1}^s {\mathcal R} [b_j] \cos (2 \pi j t/s) + {\mathcal I} [b_j]
\sin (2 \pi j t/s).
\]


