---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 5-8: Recursive Computation of Autocovariance

- A second technique finds the autocovariances without first finding the MA representation.

## Paradigm 5.8.1. Method 2 for ARMA Autocovariances 
 
- Determine a recursive relation for the $\gamma (h)$:
\[
  \gamma (k) - \sum_{j=1}^p \phi_j \gamma (k-j) = \begin{cases} \sigma^2 \sum_{j=0}^{q-k} \theta_{j+k} \psi_j \quad \mbox{if} \; k \leq q \\
  0 \qquad \mbox{if} \; k > q.
  \end{cases}
  \]
- This is compactly written as $\phi (B) \gamma_k = 0$ for $k > q$, an ODE in terms of the autocovariance
function.
- To solve, we find the roots of $\phi (z)$ and determine the homogeneous solution, using initial 
conditions for $\gamma_k$.
- If the roots $\zeta_j$ of $\phi (z)$ are distinct, then
\[
 \gamma (k) = \sum_{j=1}^p b_j \zeta_j^{-k}
\]
for coefficients $b_j$.
- These initial conditions can be recursively determined, using other expressions for
${\mathbb E} [ W_t X_{t-h}]$, where $W_t = \phi(B) X_t$ (this is the "moving average"
portion of the ARMA process).

## Proposition 5.8.3. Exponential Decay of ARMA ACF.

Consider a stationary ARMA($p$,$q$) process such that $\phi (B) X_t = \theta (B) Z_t$,
for $Z_t \sim \mbox{WN} (0, \sigma^2)$. Assume $\phi$ and $\theta$ have no common roots.
Then there exists a constant $C>0$ and $r \in (0,1)$ such that
\[
 |\gamma (k) | \leq C r^{|k|}
\]
 for all $|k| \geq \max \{ p, q+1 \}$. Hence the ACF exists.
 
## Exercise 5.51. Direct Algorithm for Autocovariance Function for the ARMA($p$,$q$)

- We encode the second method and run on Example 5.5.7.
- This encoding is *ARMAauto.r*. Most of the code has to do with computing the initial
values of the autocovariance, and the latter part of the code has the recursion.

```{r}
polymult <- function(a,b) {
bb <- c(b,rep(0,length(a)-1))
B <- toeplitz(bb)
B[lower.tri(B)] <- 0
aa <- rev(c(a,rep(0,length(b)-1)))
prod <- B %*% matrix(aa,length(aa),1)
return(rev(prod[,1]))
}

ARMAauto <- function(phi,theta,maxlag)
{
	p <- length(phi)
	q <- length(theta)
	gamMA <- polymult(c(1,theta),rev(c(1,theta)))
	gamMA <- gamMA[(q+1):(2*q+1)]
	if (p > 0) 
	{
		Amat <- matrix(0,nrow=(p+1),ncol=(2*p+1))
		for(i in 1:(p+1))
		{
			Amat[i,i:(i+p)] <- c(-1*rev(phi),1)
		}
		Amat <- cbind(Amat[,(p+1)],as.matrix(Amat[,(p+2):(2*p+1)]) +
			t(matrix(apply(t(matrix(Amat[,1:p],p+1,p)),2,rev),p,p+1)))
		Bmat <- matrix(0,nrow=(q+1),ncol=(p+q+1))
		for(i in 1:(q+1))
		{
			Bmat[i,i:(i+p)] <- c(-1*rev(phi),1)
		}
		Bmat <- t(matrix(apply(t(Bmat),2,rev),p+q+1,q+1))
		Bmat <- matrix(apply(Bmat,2,rev),q+1,p+q+1)
		Bmat <- Bmat[,1:(q+1)]
		Binv <- solve(Bmat)
		gamMix <- Binv %*% gamMA
		if (p <= q) { gamMix <- matrix(gamMix[1:(p+1),],p+1,1) 
			} else gamMix <- matrix(c(gamMix,rep(0,(p-q))),p+1,1)
		gamARMA <- solve(Amat) %*% gamMix 
	} else gamARMA <- gamMA[1]

	gamMA <- as.vector(gamMA)
	if (maxlag <= q) gamMA <- gamMA[1:(maxlag+1)] else gamMA <- c(gamMA,rep(0,(maxlag-q)))
	gamARMA <- as.vector(gamARMA)
	if (maxlag <= p) gamARMA <- gamARMA[1:(maxlag+1)] else {
	for(k in 1:(maxlag-p))
	{
		len <- length(gamARMA)
		acf <- gamMA[p+1+k]
		if (p > 0) acf <- acf + sum(phi*rev(gamARMA[(len-p+1):len]))
		gamARMA <- c(gamARMA,acf)
	} }
	return(gamARMA)
}
```

- We illustrate with particular settings.

```{r}
phi1 <- .5
theta1 <- 5/6
theta2 <- 1/6
sigma <- 1
n <- 10
my.acf <- ARMAauto(phi1,c(theta1,theta2),n)*sigma^2
plot(ts(my.acf,start=0),xlab="Lag",ylab="Autocovariance",
     ylim=c(min(my.acf),max(my.acf)),type="h")
```


