---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 2-3: Stationarity

- We want to generalize the concept of *identical distribution* to a stochastic process.

## Definition 2.3.1. Marginal Distributions

- First marginals are just the $X_t$ random variables' distributions.
- Second marginals are joint distributions for all pairs $(X_t,X_s)$.
- Third marginals are joint distributions for all triplets, etc.

## Fact 2.3.4. The First Marginal

- Saying $\{ X_t \}$ has same first marginal is same as saying they are identically distributed.
- Sometimes called *First Order Stationary*.
- In particular, all means are the same: ${\mathbf E} [X_t] = {\mathbf E} [X_s]$ for all $t,s$.

## Remark 2.3.7. Second Marginals Under Shift

- Suppose all pairs have the same distribution when shifted: 
\[
  (X_1, X_2) \sim (X_2, X_3) \sim (X_3, X_4) \ldots
\]
- Then second marginal distribution only depends on lag $h$, i.e., distribution of $(X_t, X_{t-h})$ does not depend on $t$.
- Sometimes called *Second Order Stationary*.
- Then the product mean (the covariance) depends only on lag:
\[
 {\mathbf E} [ X_t \, X_{t-h}]. 
\]
It does not depend on $t$.  
 
## Exercise 2.28. Visualizing Stationarity

- We generate 100 simulations of a Gaussian AR(1), and generate a scatterplot of $(X_1, X_2)$ 
- We repeat with $(X_3, X_4)$
 
```{r}
x1 <- NULL
x2 <- NULL
x3 <- NULL
x4 <- NULL

for(i in 1:100) {

z <- rnorm(10)
x <- rep(0,10)
phi <- .9
x0 <- rnorm(1)/sqrt(1-phi^2)
x[1] <- phi*x0 + z[1]
for(t in 2:10) { x[t] <- phi*x[t-1] + z[t] }

x1 <- c(x1,x[1])
x2 <- c(x2,x[2])
x3 <- c(x3,x[3])
x4 <- c(x4,x[4])
}

plot(x2,x1,xlab="X Past",ylab="X Present")

plot(x4,x3,xlab="X Past",ylab="X Present")
```
