---
title: 'Time Series: A First Course with Bootstrap Sampler'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/TimeSeriesR/Data')
```

# Lesson 2: Cycles and Transforms

- High association can occur for non-adjacent random variables. 
- For some fixed $h > 1$, we may have $X_t$ and $X_{t-h}$ associated for all $t$.
- This is a *periodic* effect, called a *cycle*, of period $h$.

## Example 1.2.1. Sunspots

- Wolfer sunspots series measures number of sunspots recorded each month.
- Cycles are roughly 11 years (so $h \approx 132$).

```{r}
wolfer <- read.table("wolfer.dat")
wolfer <- ts(wolfer,start=1749,frequency=12)
plot(wolfer,xlab="Year",ylab="Sunspots")
```

## Example 1.2.2. Unemployment Insurance Claims

- Weekly measurements of claims for unemployment insurance (pre-Covid).
- There is a weekly cylical pattern, corresponding to a cycle of annual period ($h=52$).

```{r}
ui <- read.table("ui.dat")
ui <- ts(ui,start=1967,frequency=52)
plot(ui,ylab="UI Claims",xlab="Year")
```

## Example 1.2.3. Mauna Loa carbon Dioxide

- Monthly measurements of CO2 levels on mount Mauna Loa.
- Apparent upward trend and monthly ($h=12$) cycle.

```{r}
mau <- read.table("mauna.dat",header=TRUE,sep="")
mau <- ts(mau,start=1958,frequency=12)
plot(mau,ylab="Mauna Loa CO2",xlab="Year")
```

## Example 1.2.4. Retail Sales of Motor Vehicles and Parts Dealers

- Monthly measurements of retail sales.
- Shows trend, monthly ($h=12$) cycle, and Great Recession.

```{r}
Ret441 <- read.table("retail441.b1",header=TRUE,skip=2)[,2]
Ret441 <- ts(Ret441,start = 1992,frequency=12)
plot(Ret441, ylab="Motor Retail Sales",xlab="Year")
```

## Example 1.2.5. Housing Starts

- Monthly measurements of housing construction started (West Region).
- Shows trend, monthly ($h=12$) cycle, and some recessions/expansions.

```{r}
Wstarts <- read.table("Wstarts.b1",header=TRUE,skip=2)[,2]
Wstarts <- ts(Wstarts,start = 1964,frequency=12)
plot(Wstarts, ylab="West Starts",xlab="Year")
```

## Log Transformation

- To visualize and model time series better, sometimes we apply a log transform (if the data is positive).
- If cycle amplitude depends on trend level, applying a log may separate this effect so that cycle amplitude is no longer growing.

```{r}
Ret441_trans <- ts(cbind(Ret441,log(Ret441)),start = 1992,frequency=12)
plot(Ret441_trans, ylab="Motor Retail Sales",xlab="Year",main="")
```

