---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/Documents/GitHub/BEwebinar/Data')
```

# Lesson 3: Time Series Regression and Autoregression

- It is tempting to regress time series data on time $t=1, \ldots, n$, as a covariate. Rarely does this provide satisfactory results (when used alone). 

## Regression on Time Trend

- Regression model with time trend:
\[
 X_t = \beta_0 + \beta_1 t + Z_t.
\]
 Is $\{ Z_t \}$ i.i.d.? Usually: **No**.

### Example 1.1.3. U.S. Population

- Try  out time trend regression for U.S. population.

```{r}
pop <- read.table("USpop.dat")
pop <- ts(pop, start = 1901)
n <- length(pop)
time <- seq(1,n)
model1 <- lm(pop ~ time)
summary(model1)
plot(ts(model1$residuals))
```

- Highly structured residuals! We could add higher order polynomial effects to time trend, but it won't really help.

## Regression on Past of Self

- Let past values of the time series be the covariates. Called *Autoregression*.
- Autoregressive model:
\[
  X_t = \rho \, X_{t-1} + Z_t.
\]
Assume $\{ Z_t \}$ i.i.d. 

### Example 1.1.3. U.S. Population

- Try  out autoregression for U.S. population.
- Here we include a constant regressor as well.

```{r}
model2 <- lm(pop[-1] ~ pop[-n])
summary(model2)
plot(ts(model2$residuals))
```

- Residuals are less structured, though maybe still not i.i.d.
- Why does this seem to work?

```{r}
cor(pop[-1],pop[-n])
plot(pop[-1],pop[-n],xlab="X Past",ylab="X Present")
```

## Incorporate Time Trend

- We can incorporate time trend into an autoregression.
- One way to do it:
\[
 X_t = \rho X_{t-1} + \beta_0 + \beta_1 t + Z_t,
\]
- This makes the mean ${\mathbf E} [ X_t ]$ depend on $\rho$.
- Another way to do it:
\begin{align*}
 Y_t & = \beta_0 + \beta_1 t + X_t \\
  X_t & = \rho \, X_{t-1} + Z_t.
\end{align*}

```{r}
model3 <- lm(pop[-1] ~ pop[-n] + time[-1])
summary(model3)
plot(ts(model3$residuals))
```

- This implements the first approach.
- Notice slope coefficient $\beta_1$ is not significant, and residuals resemble those of the pure autoregressive model. So not much benefit to using time covariate.

