---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 6-4: Toeplitz Matrices

- We discuss a decomposition of Toeplitz matrices, with connections to the spectral density.
- This is useful for model fitting and prediction.

## Fact 6.4.2. Spectral Representation of a Symmetric Matrix

- Let $A^*$ denote the conjugate transpose of a matrix $A$.
- A matrix $A$ is Hermitian if $A^* = A$.
- A matrix $U$ is unitary if $U^{-1} = U^*$.
- For any Hermitian $A$, there exists unitary $U$ such that $A = U D U^*$, where
$D$ is diagonal with real entries. 
- The columns of $U$ are the eigenvectors of $A$, and $D$ has the eigenvalues of $A$.

## Definition 6.4.3. Fourier Frequencies

- For any $n$, the *Fourier frequencies* are defined as $\lambda_{\ell} = 2 \pi \ell/n$,
for $[n/2] - n+1 \leq \ell \leq [n/2]$.
- When $n$ is even, this excludes the frequencies $\pi$ and $-\pi$ both being in the 
set, since these are redundant.
- We define an $n \times n$-dimensional matrix $Q$, whose entries are Fourier frequencies:
\[
 Q_{jk} = n^{-1/2} e^{i j \lambda_{[n/2] - n + k}}.
\]
- The matrix $Q$ is unitary, and can be used in an approximation result for Toeplitz matrices.

## Theorem 6.4.5. Spectral Decomposition of Toeplitz Covariance Matrices

- Let $\Gamma_n$ denote the autocovariance matrix of a sample of size $n$ from a 
stationary time series. Suppose the autocovariances $\gamma (k)$ are absolutely
summable. Then
\[
  \Gamma_n \approx Q \Lambda Q^*,
\]
 with $\Lambda$ diagonal with entries $f(\lambda_{[n/2] - n +k})$. The approximation 
 $\approx$ means that the difference entry-by-entry tends to zero as $n$ tends to
 $\infty$.
- It can also be shown that $\Lambda \approx Q^* \Gamma_n Q$.
- If the process is invertible, then $\Gamma_n^{-1} \approx Q \Lambda Q^*$ as well.

## Remark 6.4.11. Positive Spectral Density

Since the eigenvalues of a symmetric non-negative definite matrix $\Gamma_n$ are real 
and non-negative, we can show that the spectral density of a stationary process 
must be non-negative. 

## Exercise 6.43. Eigenvalues of an MA(1) Toeplitz Matrix.

- Consider an MA(1) with parameter $\theta = .8$.
- Compute the eigenvalues of $\Gamma_n$ for various $n$, and compare to the spectral
density evaluated at the Fourier frequencies.

```{r}

theta <- .8

n <- 10
lambda <- 2*pi*seq(0,n-1)/n
Gamma <- toeplitz(c(1+theta^2, theta, rep(0,n-2)))
eigen(Gamma)$values
rev(sort(1+theta^2 + 2*theta*cos(lambda)))

n <- 20
lambda <- 2*pi*seq(0,n-1)/n
Gamma <- toeplitz(c(1+theta^2, theta, rep(0,n-2)))
eigen(Gamma)$values
rev(sort(1+theta^2 + 2*theta*cos(lambda)))

n <- 30
lambda <- 2*pi*seq(0,n-1)/n
Gamma <- toeplitz(c(1+theta^2, theta, rep(0,n-2)))
eigen(Gamma)$values
rev(sort(1+theta^2 + 2*theta*cos(lambda)))
```

## Exercise 6.45. Eigenvalues of an MA(1) Inverse Toeplitz Matrix.

- Consider an MA(1) with parameter $\theta = .8$.
- Compute the eigenvalues of $\Gamma_n^{-1}$ for various $n$, and compare to the 
recipocal spectral density evaluated at the Fourier frequencies.

```{r}
theta <- .8

n <- 10
lambda <- 2*pi*seq(0,n-1)/n
Gamma <- toeplitz(c(1+theta^2, theta, rep(0,n-2)))
eigen(solve(Gamma))$values
1/sort(1+theta^2 + 2*theta*cos(lambda))

n <- 20
lambda <- 2*pi*seq(0,n-1)/n
Gamma <- toeplitz(c(1+theta^2, theta, rep(0,n-2)))
eigen(solve(Gamma))$values
1/sort(1+theta^2 + 2*theta*cos(lambda))

n <- 30
lambda <- 2*pi*seq(0,n-1)/n
Gamma <- toeplitz(c(1+theta^2, theta, rep(0,n-2)))
eigen(solve(Gamma))$values
1/sort(1+theta^2 + 2*theta*cos(lambda))
``` 



