---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 1-1: Time Series Data

- A *time series* is a dataset where the observations are recorded at discrete time intervals.
- We denote the obervation a time $t$ by $x_t$, and the random variable by $X_t$.
- We have times $t = 1, 2, \ldots, n$ for the *sample* $X_1, X_2, \ldots, X_n$.
- Time series data might not be i.i.d. (Independent and Identically Distributed)!

## Example 1.1.3. U.S. Population

- U.S. Population growth over the twentieth century.

```{r}
pop <- read.table("USpop.dat")
pop <- ts(pop, start = 1901)
plot(pop*10e-7,xlab="Year",ylab="U.S. Population",lwd=1)
```

## Example 1.1.4. Urban World Population

- Urban World Population trends upwards more strongly after WWII.

```{r}
urban <- read.table("urbanpop.dat")
urban <- ts(urban[67:1,], start = 1951)
plot(urban*1e-9,xlab="Year",ylab="Urban Population",lwd=1)
```

## Example 1.1.5. Non-Defense Capitalization

- Non-Defense Capitalization (New Orders) shows non-monotonic trend, and is more *noisy*.

```{r}
ndc <- read.table("Nondefcap.dat")
ndc <- ts(ndc[,2],start=c(1992,3),frequency=12,names= "NewOrders")
plot(ndc,xlab="Year",ylab="Non-defense Cap")
```

## Serial Dependence and Forecasting

- Smoothness corresponds to high positive association (correlation) between adjacent variables.
- High association means forecasting (prediction of future values) is easier.

## Example 1.1.8. Dow Jones Industrial Average

- Dow Jones Industrial Average has trend, but is hard to forecast one day ahead.

```{r}
dow <- read.table("dow.dat")
dow <- ts(dow,start=c(2008,164),frequency=252)
plot(dow,xlab="Year",ylab="Dow Levels")
```

- We can plot the log returns (consecutive difference of logged data), which shows how volatility is not constant.

```{r}
dow.diff <- diff(log(dow[,1]))
plot(dow.diff,xlab="Year",ylab="Dow Log Returns")
```