---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 6-1: Spectral Density

- We define the spectral density, which allows us to do time series analysis in
the frequency (or Fourier) domain.

## Definition 6.1.2.

- The *spectral density* of a stationary time series is
\[
 f(\lambda) = \sum_{k = -\infty}^{\infty} \gamma (k) e^{-i \lambda k},
\]
for $\lambda \in [-\pi, \pi]$.
- Also the spectral density is the restriction of the AGF to the unit circle: $f(\lambda) = G (e^{-i \lambda})$.
- A sufficient condition for existence is absolute summability of the autocovariances. 
This also guarantees $f(\lambda)$ is continuous.

## Remark 6.1.6. Spectral Representation of the Autocovariance.

- By Fourier inversion, we can recover the autocovariances from the spectral density:
\[
 \gamma (k) = \frac{1}{2 \pi} \int_{-\pi}^{\pi} f(\lambda) e^{i \lambda k} \, d\lambda.
\]
- So for $k=0$, we see the process' variance is the average integral of $f$.

## Exercise 6.4. MA(1) Spectral Density.

- For an MA(1), the spectral density is 
\[
f(\lambda) = \gamma (0) + 2 \gamma(1) \cos(\lambda) = \gamma (0) \,
\left( 1 + 2 \rho(1) \cos (\lambda) \right).
\]
- We plot $f(\lambda)$ for $\rho(1) = .4$ and $\gamma(0) =1$.
- The units are in $\pi$, called "Cycles".

```{r}
mesh <- 1000
lambda <- pi*seq(-mesh,mesh)/mesh
rho.1 <- .4
spec <- 1 + 2*rho.1*cos(lambda)
plot(ts(spec,start=-1,frequency=mesh),xlab="Cycles",ylab="")
abline(h=0,col=2)
```

## Fact 6.1.8. Further Properties of the Spectral Density

- Because $\gamma (k) = \gamma (-k)$, $f(\lambda)$ is real and even.
- Also $f(\lambda) \geq 0$ follows from positive definite property.

## Exercise 6.7. MA($q$) Spectral Density Computation.

- For an MA($q$), the spectral density is
\[
 f(\lambda) = \gamma (0) \,
\left( 1 + 2 \sum_{k=1}^q \rho(k) \cos (\lambda k) \right).
\]
- Since it is even, we can just focus on $\lambda \in [0, \pi]$.

```{r}
maq.spec <- function(ma.acf,mesh)
{
	q <- length(ma.acf)-1
	lambda <- pi*seq(0,mesh)/mesh
	spec <- ma.acf[1]*cos(0*lambda)
	if(q > 0)
	{
		for(k in 1:q)
		{
			spec <- spec + 2*ma.acf[k+1]*cos(k*lambda)
		}
	}
	return(spec)
}
```

- We plot $f(\lambda)$ with $q=2$, $\gamma(0)=1$, $\rho(1)= 0$, and $\rho(2) =.6$.
- However, these values do not correspond to a positive definite autocovariance,
and the resulting function takes negative values.

```{r}
spec <- maq.spec(c(1,0,.6),mesh)
plot(ts(spec,start=0,frequency=mesh),xlab="Cycles",ylab="")
abline(h=0,col=2)
```

## Corollary 6.1.9.

- Suppose we filter stationary $\{ X_t \}$ with some $\psi (B)$, yielding $Y_t = \psi (B) X_t$. 
Then the spectral densities of input and output are related by
\[
   f_y (\lambda) = {| \psi (e^{-i \lambda}) |}^2  f_x (\lambda).
\]
- This follows from Theorem 5.6.6.
- We call $\psi(e^{-i \lambda})$ the *frequency response function* of the filter $\psi(B)$.
- We call ${|\psi(e^{-i \lambda}) |}^2$ the *squared gain function* of the filter $\psi(B)$.

## Theorem 6.1.12.

Let $\{ X_t \}$ be a stationary ARMA($p$,$q$) process such that $\phi (B) X_t = \theta (B) Z_t$,
for $Z_t \sim \mbox{WN} (0, \sigma^2)$. Suppose $\phi (z)$ has no roots on the unit
circle. Then the spectral density exists:
\[
 f (\lambda) = \sigma^2 \, \frac{ {|\theta (e^{-i \lambda}) |}^2 }{ {|\phi (e^{-i \lambda}) |}^2  }
\]

## Exercise 6.12. ARMA($p$,$q$) Spectral Density. 

- We write code for the ARMA spectral density, based on the formula of Theorem 6.1.12, 
taking as input the $\theta (z)$ and $\phi (z)$ polynomials.

```{r}
armapq.spec <- function(ar.coef,ma.coef,sigma,mesh)
{
	p <- length(ar.coef)
	q <- length(ma.coef)
	lambda <- pi*seq(0,mesh)/mesh
	spec.ar <- rep(1,mesh+1) 
	if(p > 0)
	{
		for(k in 1:p)
		{
			spec.ar <- spec.ar - ar.coef[k]*exp(-1i*lambda*k)
		}
	}
	spec.ma <- rep(1,mesh+1) 
	if(q > 0)
	{
		for(k in 1:q)
		{
			spec.ma <- spec.ma + ma.coef[k]*exp(-1i*lambda*k)
		}
	}
	spec <- sigma^2*Mod(spec.ma)^2/Mod(spec.ar)^2
	return(spec)
}
```

- We plot the spectral density of the cyclic ARMA(2,1) process of Example 5.7.2:
for $\rho \in (0,1)$ and $\omega \in (0,\pi)$, let $\{ X_t \}$ satisfy
\[
 ( 1 - 2 \rho \cos (\omega) B + \rho^2 B^2) X_t = (1 - \rho \cos (\omega) B) Z_t.
\]
- We set $\rho = .8$ and $\omega = \pi /6$.

```{r}
spec <- NULL
mesh <- 1000
rho <- .8
omega <- pi/6
ar.coef <- c(2*rho*cos(omega),-1*rho^2)
ma.coef <- -1*rho*cos(omega)
spec <- armapq.spec(ar.coef,ma.coef,1,mesh)
plot(ts(spec,start=0,frequency=mesh),xlab="Cycles",ylab="",main="")
```
   
## Corollary 6.1.14. 

- Let $\{ X_t \}$ be a weakly stationary, mean zero time series with strictly positive
spectral density of form given in Theorem 6.1.12. Then there exists a white noise
$\{ Z_t \}$ such that $\phi (B) X_t = \theta (B) Z_t$.
- This is proved by defining $Z_t = \psi (B) X_t$ with $\psi (z) = \phi (z)/\theta (z)$,
and checking that $\{ Z_t \}$ is white noise. 
- This $\psi (B)$ is a *whitening filter*. It transforms a time series to white noise!

## Theorem 6.1.16. MA($\infty$) Representation.

Let $\{ X_t \}$ be a weakly stationary, mean zero time series with autocovariance
function $\gamma (k)$ that is absolutely summable, and positive spectral density.
Then $\{ X_t \}$ is an MA($\infty$) process with respect to some white noise $\{ Z_t \}$:
\[
 X_t = \sum_{j \geq 0} \psi_j Z_{t-j},
\]
and $\psi_0 =1$.

## Corollary 6.1.17. AR($\infty$) Representation.

Under the assumptions of Theorem 6.1.16, $\{ X_t \}$ is an AR($\infty$) process with
respect to the same white noise $\{ Z_t \}$:
\[
 X_t = - \sum_{j \geq 1} \pi_j  X_{t-j} + Z_t.
\]
