---
title: 'Time Series: A First Course with Bootstrap Starter'
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/OneDrive/Documents/GitHub/BEwebinar/Data')
```

# Lesson 10-3: AR Identification

## Remark 10.3.1. AR Model Identification

- To identify an AR model, we must use the sample PACF rather than the true PACF.
- Take as null hypothesis that the process is an AR(p), such that
\[
 H_0 :  \kappa (h) = 0 \quad \mbox{for all} \;  h > p.
\]
- We can use the estimate $\widehat{\kappa} (h)$ as test statistic, for any $h > p$.
- We require an asymptotic distribution theory.
   

## Corollary 10.3.4.

Suppose $\{ X_t \}$ is an AR($p$) process with mean $\mu$ and i.i.d. inputs of variance $\sigma^2$. Then
\[
  \sqrt{n}  (\widehat{\kappa} (h) - \kappa (h)) \Rightarrow \mathcal{N} (0, \sigma^2
    \underline{e}_h^{\prime} \Gamma_h^{-1} \underline{e}_h ),
\]
 where $\underline{e}_h$ is the unit vector with one in last (index $h$) position.


## Paradigm 10.3.6. AR Model Identification

- Suppose the true model is AR(p), and we test $H_0$ with $h > p$.
- Trick: the process is also an AR(h), where the last $h-p$ coefficients are zero!
- So we can apply Corollary 10.3.4, and it can be shown that because $h > p$ the limiting variance is $1$, i.e.,
\[
   \sqrt{n} \, \widehat{\kappa} \Rightarrow \mathcal{N} (0,1)
\]
under $H_0$.  
- So we can do sequential testing using the sample PACF, with the same cautions about multiple testing.
- We also have an *Empirical Rule* for AR identification.
- **Empirical Rule**: let $\widehat{p}$ be the smallest positive integer such that 
\[
 | \widehat{\kappa} (\widehat{p}+k)| < c \sqrt{n^{-1} \log n}
\]
 for all $k=1, \ldots, K_n$, where $c > 0$ is a fixed constant, and $K_n$ is a positive, non-decreasing integer-valued function of $n$ such that $K_n = o(\log n)$.
 
HERE: maybe eliminate this example? 
 
## Example of U.S. Population

- For the U.S. Population time series, we difference once and then plot the sample ACF and PACF. Dotted blue lines are confidence limits (discussed below).
- We should guess an AR(1) model.

```{r}
pop <- read.table("USpop.dat")
pop <- ts(pop, start = 1901)
acf(diff(pop))
pacf(diff(pop))
```
 
HERE: change to Example 10.3.7???

## Example of Identifying AR Order for U.S. Population

- For the U.S. Population time series, we difference once and compute sample PACF. 
- Dotted blue lines are confidence limits, based on asymptotic variance of $1/n$, and quantiles from the normal at a 5% level.
- The AR(1) specification seems reasonable, although the lag 8 test statistic is barely significant.

```{r}
n <- length(diff(pop))
pop.pacf <- pacf(diff(pop))
tstat <- sqrt(n)*pop.pacf$acf[1:8]
round(rbind(tstat,rep(qnorm(1-.05/2),8)),digits=5)
```

- We also apply the *Empirical Rule*.
 
```{r}
alpha <- .05
crit <- qnorm(1-alpha/2)
K.n <- 1 + floor(3*sqrt(log(n,base=10)))
kappa.test <- pop.pacf$acf/sqrt(log(n,base=10)/n)
k <- 1
while(k < (n-K.n))
{
	if(max(abs(kappa.test[k:(k+K.n-1)])) < crit) 
	{ 
		p.hat <- k-1
		k <- n-K.n
	} else { k <- k+1 }
}
```

- This results in AR identification of order `r p.hat`.

